ScPy: A SuperCollider Extension for Performing Numerical Computation via Embedded Python

### 2016-08-28 doc/paper.tex
more literature

\begin{figure}[ht]
    \caption{NIME dimension space analysis for ScPy.}\label{fig:nime}
    \begin{center}
        \begin{tikzpicture}[rotate=-116,xscale=-1]
            \tkzKiviatDiagram[label distance=.5cm,
                    radial=7,
                    gap=0.75,
                    lattice=4]{Required Expertise,
                    Musical Control,
                    Degrees of Freedom (Inputs),
                    Feedback Modalities (Outputs),
                    Inter-actors,
                    Distribution in Space,
                    Role of Sound}
            \tkzKiviatLine[thick,color=blue,mark=none,
                           fill=blue!20,opacity=.5](4,4,1.5,1.5,0.5,0.5,3)
        \end{tikzpicture}
    \end{center}
\end{figure}

@@

ScPy can be used for live coding of audio effects, similarly to how the environments studied in
this paper can be used to live code synths and pattens. However, at the time of writing one issue
is blocking ScPy from realistically being used for live coding performances: the Python global
environment can be put into a state which prevents the code from executing without throwing an
exception, which could halt a performance in its tracks if the programmer made one of a number of
errors. This may be avoidable if more advanced error detection is implemented in the future.

@@

@@

Although designing a domain-specific language (DSL)~\cite{van2000domain} for this task was briefly
considered, we eventually decided to use Python for a number of reasons. First, Python includes an
excellent C API which makes embedding it within other languages trivial in comparison to many other
options. Second, Python's syntax is easily readable and concise --- it clearly adheres to our
goals. Finally, Python has a massive number of community built libraries for performant scientific
and numerical computing. With a few small exceptions, which are clearly specific to spectral data
(such as the phase vocoder), these libraries contained every one of the fundamental operations we
could think of.

@@

In order to facilitate the use of ScPy in live coding performances, many errors are ignored or
handled in a way that can be easily recovered from. In particular, floating point errors such as
division by zero or other invalid operations are ignored. Special values such as infinity or NAN
are replaced with zeros. This is done because SuperCollider has poor handling of these special
values and in some cases will crash if they are found. Python exceptions are printed to the
SuperCollider post window, where they can easily be examined and corrected by the user.

>>>>

\begin{figure}[ht]
    \caption{NIME dimension space analysis for ScPy.}\label{fig:nime}
    \begin{center}
        \begin{tikzpicture}[rotate=-116,xscale=-1]
            \tkzKiviatDiagram[label distance=.5cm,
                    radial=7,
                    gap=0.75,
                    lattice=4]{Required Expertise,
                    Musical Control,
                    Degrees of Freedom (Inputs),
                    Feedback Modalities (Outputs),
                    Inter-actors,
                    Distribution in Space,
                    Role of Sound}
            \tkzKiviatLine[thick,color=blue,mark=none,
                           fill=blue,opacity=.5](4,4,1.5,1.5,0.5,0.5,3)
        \end{tikzpicture}
    \end{center}
\end{figure}

@@

ScPy can be used for live coding of audio effects, similarly to how the environments studied in
this paper can be used to live code synths and pattens. More details about how ScPy was designed to
facilitate this can be found in \autoref{sec:methodology}.

@@

Laroche and Dolson~\cite{laroche1999improved} present an improvement to the phase vocoder which
aims to prevent the ``phasiness'' problem where different parts of a sound lose their phase
coherence, as well as provide a significant decrease in computational cost. The paper recognizes
that there are two types of phase coherence that must be preserved in any STFT based algorithm to
ensure that the reproduced sound is without artifacts: horizontal and vertical. The basic phase
vocoder algorithm is built to preserve horizontal phase coherence --- but it does so at the expense
of vertical phase coherence. To remedy this, they build on prior work by presenting two new
phase-locking techniques.

Similar techniques~\cite{laroche1999new} enable new phase vocoder techniques that enable
exotic audio effects. The core of the techniques is a peak detection process followed by peak
shifting. Since these new techniques operate on peaks, the standard phase vocoder algorithm can be
simplified to not require phase unwrapping, reducing computational complexity. Additionally, they
produce better quality output than the standard phase vocoder algorithm because of implicit
phase-locking.

The phase vocoder currently implemented in ScPy uses the standard algorithm. Future experiments
could implement many of the proposed modifications to the phase vocoder to improve sound quality.

Bradford et al.\ have published a number of works detailing the sliding
DFT~\cite{bradford2005sliding} and associated sliding phase vocoder~\cite{bradford2007sliding}.
These new algorithms are efficient implementations for the STFT with a hop size of a single sample.
Although such a small hop size would be impossible to process in real time using current hardware
and the STFT, the sliding DFT makes it possible using consumer GPU
hardware~\cite{bradford2011real}. The sliding phase vocoder additionally has a number of advantges
over traditional implementations. Quality is greatly improved (at a high computational cost). Each
analysis bin can hold any frequency, essentially making the spectrum more of a generic oscillator
bank than a traditional spectrum. This simplifies effects such as pitch shifting to a simple vector
multiplication in the frequency domain. Additionally, since the time resolution is so fine, effects
that require time modulation such as FM or AM can now be performed in the frequency domain.

Using Theano, the sliding DFT and sliding phase vocoder could be implemented easily in ScPy. It
remains unknown if it will be possible to run them in real time, but even if not this future
extension could add the possibility to experiment with many novel effects.

@@

Although designing a domain-specific language for this task was briefly considered, we eventually
decided to use Python for a number of reasons. First, Python includes an excellent C API which
makes embedding it within other languages trivial in comparison to many other options. Second,
Python's syntax is easily readable and concise --- it clearly adheres to our goals. Finally, Python
has a massive number of community built libraries for performant scientific and numerical
computing. With a few small exceptions, which are clearly specific to spectral data (such as the
phase vocoder), these libraries contained every one of the fundamental operations we could think
of.

@@

In order to facilitate the use of ScPy in live coding performances, many errors are ignored or
handled in a way that can be easily recovered from. In particular, floating point errors such as
division by zero or other invalid operations are ignored. Special values such as infinity or NaN
are replaced with zeros. This is done because SuperCollider has poor handling of these special
values and in some cases will crash if they are found. Python exceptions are printed to the
SuperCollider post window, where they can easily be examined and corrected by the user.
